EEG Waveform Analysis of P300 ERP with applications to Brain Computer Interfaces

### 2018-10-26 BrainScience/response.letter.1.tex

6. The Results section is not well-presented. The authors are advised to include a paragraph before Table 2, Table 3 and Fig. 6-8 presenting the obtained results (best/worst channel performance and algorithm performance for each experiment) and then discuss the results in the Discussion section. Please revise excessive descriptions in the captions of Table 2, Table 3 and Fig. 6-8.

7. “the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of 70\%” (Page 19, Paragraph 2, Line 465-466) What is the obtained character identification rate in other scientific studies? Is the obtained result (70\%) of this study comparable to others?

8. In the Discussion section a table presenting the statistical analysis results would present better the results and would facilitate the reader.

\subsection*{\ovalbox{General Comments}}
The paper is an interesting research paper and presents an analysis of the P300 ERP. The authors provide a literature review of the current methods used to detect EEG patterns and conduct three experiments to identify the P300 ERP in EEG recordings. The paper is well-organized. The paper should incorporate the following comments before its acceptance for publication:
\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you for your encouraging and thoughtful comments. We have modified several points according to your recommendations. We hope the article is now better suited for publication. 
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

1. The major concern is that the experiments have been conducted with only one participant. The authors should collect more data in this study. In other studies cited in the paper by authors,  two and 16 participants were recruited.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We have now included more participants to this experiment and it was divided in two modalities.  The first, which is now called passive-modality, includes now 4 subjects.  The second, an active-modality, now includes 4 more participants who were engaged in a copy-spelling task on a P300-Based BCI Speller.   For this subjects, the pseudo-real dataset is constructed by superimposed a P300 template obtained for each subject into segments marked as hit.  The baseline for each segment is constructed from a random no-hit segment for the same subject.  
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Thank you very much for your comment.  We have modified the specific paragraphs that the Reviewer kindly offered as example and we have revised the entire manuscript modifying different parts, simplifying sections.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

3. In the Introduction section the authors refer to BCI applications; however, no scientific studies are reported. Also, the literature section should be revised with recent publication. Below are some scientific studies that is suggested to be included for BCI applications:

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

Absolutely thank you very much for providing these very important references.  We added all of them.

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

4. In the literature section, Table I briefly presents the proposed methods followed by other authors; however, there is no text describing the table. The authors should include a paragraph describing each method and the recent findings concisely, in order to present better the literature review.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We have included a new paragraph describing the depictions which are shown in the table and in which areas are they used for.

@@

5. In Figure 2 please insert the y-axis title to clearly present ERP template.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Our applogize for this mistake.  We have corrected it and revised all the other figures.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

6. The Results section is not well-presented. The authors are advised to include a paragraph before Table 2, Table 3 and Fig. 6-8 presenting the obtained results (best/worst channel performance and algorithm performance for each experiment) and then discuss the results in the Discussion section. Please revise excessive descriptions in the captions of Table 2, Table 3 and Fig. 6-8.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We have included the best/worst channel obtained for each algorithm, subject and experiment. This enhancement contributes greatly to the understanding of the obtained values.  We have also modified the captions of the tables and figures, thank you very much for raising this issue.
With this, we hope we have full addressed your point.
 }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

7. “the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of 70\%” (Page 19, Paragraph 2, Line 465-466) What is the obtained character identification rate in other scientific studies? Is the obtained result (70\%) of this study comparable to others?

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Studies on this dataset reported performance values of perfect classification.  However in this case, as we are emphasizing the waveforms of the signals, the classification uses only single-channel activity, which is less information.  Regardless of this, the obtained results for some of the algorithms can be used to implement usable P300-Based Speller applications due to the usage of word predictive schemes.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

### 2018-10-26 BrainScience/response.letter.1.tex

5. In Figure 2 please insert the y-axis title to clearly present ERP template.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
Our applogize for this mistake.  We have corrected it and revised all the other figures.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

6. The Results section is not well-presented. The authors are advised to include a paragraph before Table 2, Table 3 and Fig. 6-8 presenting the obtained results (best/worst channel performance and algorithm performance for each experiment) and then discuss the results in the Discussion section. Please revise excessive descriptions in the captions of Table 2, Table 3 and Fig. 6-8.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
We have included the best/worst channel obtained for each algorithm, subject and experiment. We hope this enhancement contributes greatly to the understanding of the obtained values.  We have also modified the captions of the tables and figures, thank you very much for raising this issue.
With this, we hope we have full addressed your point.
 }
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

@@

7. “the obtained character identification rate is above theoretical chance level, and for some algorithms close to the usable threshold of 70\%” (Page 19, Paragraph 2, Line 465-466) What is the obtained character identification rate in other scientific studies? Is the obtained result (70\%) of this study comparable to others?

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
The reported binary classification accuracy for this BCI Competition was around 95\%. However in this case, as we are emphasizing the waveforms of the signals, the classification uses only single-channel activity, which is less information.  Regardless of this, the obtained results for some of the algorithms can be used to implement usable P300-Based Speller applications due to the usage of word predictive schemes.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

### 2018-10-26 BrainScience/response.letter.2.tex

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}
These important references were added to the manuscript.  Thank you for pointing them out.
}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

4. Page 12, Lines 383-394. It should be noted that the dataset from which the template ERP subject was drawn from was an ALS patient dataset. It is currently unclear why Subject Number 8 and trial number 2 was selected out of all the subjects and trials in the BNCI-Horizon database.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

The reason we selected that particular trial and subject was because our manuscript stressed the importance of the waveform, the shape of the ERP.  The obtained P300 waveform of Subject Number 8, trial 2 was the most prototypical that we've found along the BNCI-Horizon dataset of ALS patients. The analysis was performed by visual observation.  We have now added also 4 more for the Active part of the pseudo-real dataset.

@@

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

The recording conditions of both datasets are exactly the same.  The EEG device, though not the same exact model, is from the same manufacturer and the sampling frequency is as close as possible. All the preprocessing stage was implemented according to~\citep{Riccio2013}.

Following Reviewer comments, the pseudo-real dataset is now divided in two.  First, we've replicated the experiment conducted by~\cite{Riccio2013} et al on 4 healthy subjects performing an active-modality P300 copy-spelling task while using the P300-Based BCI Speller. We also extended the original single null-EEG stream to 4 subjects while performing the passive-modality where the letter to copy-spell is not informed to the user and the subject is unaware of how the matrix works.

Regarding the active-modality, for each subject, the most prototypical P300 waveform template is selected and extracted.  This template is superimposed on each preprocessed segment marked as hit.  The baseline was obtained by picking randomly any no-hit segment for the same subject.

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

6. The separate manipulation of latency, as well as amplitude noise, is interesting and if extended may in and of itself prove to be an interesting manuscript in and of itself (e.g. contrasting the impact of a range of different latency variations and amplitude noise on BCI classification ability), true EEG data almost invariantly involves not only a mix of both noise sources but other sources such as cognitive fatigue, systematic task stimuli adaptation etc. Thus all three experiments represent assessments of isolated aspects of noise which would be interesting if expanded on in a more expansive simulation set (see point 4 and summary) but currently are too limited in scope to draw strong conclusions regarding the expected inter-component performance in real-world data.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We appreciate a lot this insight which we also found interest to look after and analyze more in detail. However, we would like to clarify that the purpose of these experiments were to assess the incidence of the most common source of noise in the generation of the P300, to the shape of the obtained waveform of the P300, "how the averaged signal looked-like".

@@

}
\end{quotation}
%
\vskip+1ex
\noindent \dotfill
\vskip+1ex

It seems as though the entire study is built on the classification of a single active P300 BCI-Speller template from an ALS participant into a single null-EEG stream of a healthy subject performing a passive-viewing version of a BCI-Speller task. The generalizability of these results are thus difficult, if not impossible to assess. Given that assessments of classification methods for P300 BCI interfaces are fairly common (e.g. Manyakov et al., 2011; Oweis et al. 2013; Krusienski et al. 2010), it is unclear what the value of another contrast of classification methods is to the field. The assessment of the impact of sources of noise is interesting (see point 6), but would need to be expanded on considerably to be of practical value. Furthermore, while this manuscript frequently emphasizes the value of BCI to the clinical field. However, most clinicians are generally uninterested in classifying the presence/absence of a P300 response, arguably the most obvious cognitive ERP component extant. Instead, the focus in that area is instead on subtler dynamics in the EEG signal such as frequency slowing, triphasic waves, and generalized ictal activity. Indeed, the introduction of this manuscript mentions several EEG components of value which notably does not include the P300 in its list. If demonstrating clinical utility is of interest, it would seem more appropriate to target the classification of these subtler elements with a larger sample set and broader response ERP pool.

\vskip+1ex
\noindent \dotfill
\vskip+1ex
%
\begin{quotation}
{\color{blue}

We would like to emphasize that the purpose of this work was to raise awareness about the automatic procedures that can analyze EEG signals by their waveforms, mimicking what traditionally EEG physicians performed on the clinical EEG.  As far as we are aware of, no work has addressed this topic in this form and we humbly believe it could be a contribution.  The Reviewer is right, but the objective of this manuscript was to establish the connection, and to focus on the goal of the BCI discipline to foster interaction between the clinical and BCI community~\citep{Chavarriaga2017}.

### 2018-10-26 BrainScience/brainscience.tex
Upload a new revised version

\begin{table}[H]
\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain. Additionally, the control algorithm SVM Support Vector Machines is included for comparison.  All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown. In this case with absence of null-signals, it can be interpreted as the channel that adds less noise to the ERP template.  All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal. }
\centering
%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
\begin{tabular}{ccccc}
\toprule
\textbf{Method}	& \textbf{Channel} &   \multicolumn{3}{c}{Performance} \\
%\cline{3-5} \\
 	&  &  \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}\\
\midrule
MP 1 & PO8  & $67\%$ & $15\%$ & $50\%$\\
MP 2 & PO7 & $24\%$ & $6\%$ & $10\%$\\
SIFT  & PO8 & $91\%$ & $18\%$ & $66\%$\\
PE     & Cz & $61\%$ & $9\%$ & $32\%$\\
SHCC & P4 & $98\%$ & $31\%$ & $80\%$\\
SVM     & PO8  & $78\%$ & $7\%$ & $53\%$\\
\bottomrule
\end{tabular}
\label{tab:results}
\end{table}

@@

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/CrossPerformanceTest.eps}
\caption{Speller performance obtained for each method for the Experiment 1.  Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
\label{fig:performancetest}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/CrossPerformanceTestLatency.eps}
\caption{Speller performance obtained for each method while latencies are artificially added to each single-intensification segment corresponding to the Experiment 2.  The achieved performance is significantly reduced for all methods. Y-axis shows letter identification performance while X-axis shows the number of intensification sequences used to calculate the ensemble average.}
\label{fig:performancetestlatency}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=15cm]{images/CrossPerformanceTestAmplitude.eps}
\caption{Speller performance obtained for the Experiment 3 while the amplitudes of the P3b component of the superimposed ERP is randomly reduced. Y-axis shows performance accuracy while X-axis shows the number of intensification sequences.}
\label{fig:performancetestamplitude}
\end{figure}

>>>>

%\begin{table}[H]
%\caption{Speller classification performance obtained for all the waveform-based algorithms: MP Matching Pursuit, SIFT Scale Invariant Feature Transform, PE Permutation Entropy and SHCC Slope Horizontal Code Chain. Additionally, the control algorithm SVM Support Vector Machines is included for comparison.  All the methods process the signal on a channel-by-channel basis, hence the best performing channel is also shown. In this case with absence of null-signals, it can be interpreted as the channel that adds less noise to the ERP template.  All the methods used $10$ intensification sequences to coherently average the trials to obtain the averaged signal. }
%\centering
%%% \tablesize{} %% You can specify the fontsize here, e.g.  \tablesize{\footnotesize}. If commented out \small will be used.
%\begin{tabular}{c|cccc|cccc|cccc|cccc|cccc|cccc|cccc|cccc}
%\toprule
%\textbf{Method}	& \multicolumn{4}{c}{Subject 1} & \multicolumn{4}{c}{Subject 2} & \multicolumn{4}{c}{Subject 3} & \multicolumn{4}{c}{Subject 4} & \multicolumn{4}{c}{Subject 5} & \multicolumn{4}{c}{Subject 6} & \multicolumn{4}{c}{Subject 7} & \multicolumn{4}{c}{Subject 8} \\
%%\cline{3-5} \\
% 	&  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}  &  \textbf{Channel} & \textbf{Experiment 1} & \textbf{Experiment 2}	& \textbf{Experiment 3}\\
%\midrule
%MP 1 & PO8  & $67\%$ & $15\%$ & $50\%$  & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$   \\
%MP 2 & PO7 & $24\%$ & $6\%$ & $10\%$  & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
%SIFT  & PO8 & $91\%$ & $18\%$ & $66\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
%PE     & Cz & $61\%$ & $9\%$ & $32\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
%SHCC & P4 & $98\%$ & $31\%$ & $80\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
%SVM     & PO8  & $78\%$ & $7\%$ & $53\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ & PO8  & $67\%$ & $15\%$ & $50\%$ \\
%\bottomrule
%\end{tabular}
%\label{tab:results}
%\end{table}

@@

\begin{figure}[H]
\centering
\includegraphics[width=7.5cm]{images/1-1.eps}
\includegraphics[width=7.5cm]{images/2-1.eps}
\caption{Speller performance obtained for each method for the Experiment 1.  Y-axis shows performance accuracy while X-axis shows the number of intensification sequences used to calculate the point-to-point signal average.}
\label{fig:performancetest}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=7.5cm]{images/1-2.eps}
\includegraphics[width=7.5cm]{images/2-2.eps}
\caption{Speller performance obtained for each method while latencies are artificially added to each single-intensification segment corresponding to the Experiment 2.  The achieved performance is significantly reduced for all methods. Y-axis shows letter identification performance while X-axis shows the number of intensification sequences used to calculate the ensemble average.}
\label{fig:performancetestlatency}
\end{figure}

\begin{figure}[H]
\centering
\includegraphics[width=7.5cm]{images/1-3.eps}
\includegraphics[width=7.5cm]{images/2-3.eps}
\caption{Speller performance obtained for the Experiment 3 while the amplitudes of the P3b component of the superimposed ERP is randomly reduced. Y-axis shows performance accuracy while X-axis shows the number of intensification sequences.}
\label{fig:performancetestamplitude}
\end{figure}
